The competing effects of the immune system and EMT on epithelial cancers

### 2019-04-19 Paper/Paper.tex
minor changes throughout

Cancer and the immune system interact in myriad ways. Beyond direct tumor-immune cell interactions, the immune system modulates the tumor microenvironment (TME), indeed, immune cells can be considered as a part of the TME: responding to local inflammatory signals, targeting mutated cells, and eradicating tumor cells to potentially restore homeostasis in a local context \cite{de2006paradoxical}. Recent breakthroughs in the field of immunotherapy are beginning to produce therapies that have had a significant impact on patient health and survival \cite{pardoll2012blockade,restifo2012adoptive}.

@@

Tumor cells interact with the immune system by initiating a sequence of events that results in the activation of the adaptive immune response. This is accomplished by the presentation of antigens recognized by innate immune cells that are transported to lymph nodes where T cells (and other components) can be activated \cite{schreiber11_cancer}. The tumor also engages in cellular processes that indirectly shape the TME, for example by releasing transforming growth factor beta (TGF-$\beta$), thus shifting the TME towards a tumor-supportive environment by enhancing immunosuppression via Treg cell activation \cite{schreiber11_cancer}.

@@

Upon carrying out their effector functions, these cytotoxic cells lose efficacy or deactivate entirely\tcr{[REF]}.
The regulatory branch of the immune system (T regulatory cells (Tregs) and other factors), inhibits the effective functioning of the cytotoxic branch\cite{ruffell2010lymphocytes}.

@@

The other facet of mesenchymal cells is their slower proliferation rate, and this is tied to the observation that mesenchymal cells often express a more stem-like phenotype.
Indeed, even in healthy tissue, this is readily observed and seen as a useful mechanism for homeostasis as epithelial cells can transition to mesenchymal, stem-like cells under appropriate conditions \tcr{[REF]}.
While many qualities are associated with being stem-like, we choose to focus on a reduced proliferation rate\cite{woods2014effects}.
Cancer cells are well-known for achieving unregulated proliferation, so exploring a subset of cancer cells with decreased proliferation could provide intriguing new results.
In our model, we refer to this property as mesenchymal growth arrest, or MGA.

@@

Here we develop a model with the goal of studying interactions between the tumor, the immune system and EMT. We seek to describe a set of crucial molecular and cellular interactions in (epithelial) tissue cells, including effects due to DNA damage and mutation, to investigate the probability that cancer will occur and, if so, when. A recent model of cancer-immune interactions \cite{guo2017multiscale} described the effects of the TME on the risk of cancer, and we build on the core cell cycle component of this model, adding significant new interactions to the immune component of the model (which was previously modeled by a single interaction), as well as the EMT component to the model. We seek to understand whether this more complex immune module will change our understanding of inflammatory effects on the tumor, and how the epithelial-mesenchymal axis influences these. In the next section we develop the model modularly to explain the intuition behind each component, and then go on to analyze its behavior. Global sensitivity analysis identifies parameters that are crucial for tumorigenesis. We then study these in more depth, focusing on the competing effects of EMT and of the immune system on tumorigenesis: we find that EMT intricately regulates tumorigenesis and that under certain regimes a careful balance of EMT- and immune-driven processes can significantly prolong cancer-free survival. To test these predictions, we analyze data from the Cancer Genome Atlas (TCGA), and find evidence for the synergistic effects predicted by the model for patients with pancreatic cancer.

@@

%%%%%%%%%%%%%%%%%%%%%%%
%                          RESULTS                         %
%%%%%%%%%%%%%%%%%%%%%%%

\section{Results}

@@

\begin{figure}
\center
{\includegraphics[width=0.8\textwidth]{Figure1/Figure1.pdf}}
\caption{A. Schematic depiction of agent-based model components; blue/red denote tissue cells with/without mutations. Black arrows depict regulation of the cell fate in each cell cycle. Inset depicts major interactions between the immune system and tissue cells.
 %The parameters used in this model can be found in the Supplementary material. Unless otherwise stated, parameter values are assumed to be those found in the Supplement. \tcr{Thoughts on the wording here? Once I have the table, I will use the name of that table.}
B. A representative simulation of one patient. The model parameter values used can be found in Supplementary Table 2. 
The inflammation cycling scheme is represented above the patient dynamics. The vertical dashed line denotes the end of the warmup period. Mut: mutated cells; Mes: mesenchymal cells. 
C. Survival curve for one cohort of patients with the parameter values given in Supplementary Table 2. }
\label{fig:ModelIntro}
\end{figure}

@@

These results represent promising steps in understanding the competing roles of the immune system and EMT during (pre-) development of epithelial cancers, yet much remains to be done. Further development of the inflammation module of this model is important given the large and sometimes paradoxical roles that the inflammatory state exerts on the epithelia and cancer-free survival (Figs. \ref{fig:MOAT} and \ref{fig:fig:VaryINFL_and_MesPars}B-D). Currently, inflammation is modeled as independently cycling between high and low schemes, however many of the agents considered in the model actively contribute to the inflammatory state, thus it would be interesting to consider connecting these components (for example by letting the level of inflammation be conditional on the number of mutant cells in the tissue). Another layer of complexity is revealed by the natural anti-inflammatory role of Tregs. One consequence of the current model is that decreasing the number of Tregs increases the cancer-free survival. Clearly, there exists a trade-off to be accounted for, and adding to the model the main effector function of Tregs could remedy this and add depth to our understanding of the various roles that Tregs play in and around the tumor.

>>>>

Cancer and the immune system interact in myriad ways. Beyond direct tumor-immune cell interactions, the immune system modulates the tumor microenvironment (TME): responding to local inflammatory signals, targeting mutated cells, and eradicating tumor cells to potentially restore homeostasis in a local context \cite{de2006paradoxical}. Recent breakthroughs in the field of immunotherapy are beginning to produce therapies that have had a significant impact on patient health and survival \cite{pardoll2012blockade,restifo2012adoptive}.

@@

Tumor cells interact with the immune system by initiating a sequence of events that results in the activation of the adaptive immune response. This is accomplished by the presentation of antigens recognized by innate immune cells that are transported to lymph nodes where T cells (and other components) can be activated \cite{schreiber11_cancer}. The tumor also engages in cellular processes that indirectly shape the TME, for example by releasing transforming growth factor beta (TGF-$\beta$), thus shifting the TME towards a tumor-supportive environment by enhancing immunosuppression via activation of T regulatory cells (Tregs)  \cite{schreiber11_cancer}.

@@

Upon carrying out their effector functions, these cytotoxic cells lose efficacy or deactivate entirely \cite{finn12_immunooncology-1}. The regulatory branch of the immune system (Tregs and other factors), inhibits the effective functioning of the cytotoxic branch\cite{ruffell2010lymphocytes}.

@@

The other facet of mesenchymal cells is their slower proliferation rate, and this is tied to the observation that mesenchymal cells have been speculated to exhibit a more ``stemness'' phenotype.
%Indeed, even in healthy tissue, this is readily observed and seen as a useful mechanism for homeostasis as epithelial cells can transition to mesenchymal, stem-like cells under appropriate conditions \tcr{[REF]}.
While many cell behaviors are associated with stemness, here we target the specific reduction in proliferation rate  \cite{woods2014effects} (which, under certain conditions can be called quiescence, though we avoid this term as it can be controversial, and is not a necessary assumption to the model developed below).
Cancer cells are of course characterized by some degree of unregulated proliferation, thus it is appealing to explore the effects of differential proliferation rates within the cancer cell population.
In our model, we refer to the reduced proliferative capacity associated with a mesenchymal cell phenotype as mesenchymal growth arrest (MGA).

@@

Here we develop a model with the goal of studying interactions between the tumor, the immune system and EMT. We seek to describe a set of crucial molecular and cellular interactions in epithelial tissue cells, including effects due to DNA damage and mutation, to investigate the probability that cancer will occur and, if so, when. A recent model of cancer-immune interactions \cite{guo2017multiscale} described the effects of the TME on the risk of cancer, and we build on the core cell cycle component of this model, adding significant new interactions to the immune component of the model (which was previously modeled by a single interaction), as well as adding the effects of EMT. We seek to understand whether this more complex immune module will change our understanding of inflammatory effects on the tumor, and how the epithelial-mesenchymal axis influences these.
In the next section we develop the model and give explanation of the intuition behind each component. We then go on to analyze its behavior. Global sensitivity analysis identifies parameters that are crucial for tumorigenesis. We then study these in more depth, focusing on the competing effects of EMT and of the immune system on tumorigenesis: we find that EMT intricately regulates tumorigenesis and that under certain regimes a careful balance of EMT- and immune-driven processes can significantly prolong cancer-free survival. To test these predictions, we analyze data from the Cancer Genome Atlas (TCGA), and find evidence for the synergistic effects predicted by the model for patients with pancreatic cancer.

@@

%%%%%%%%%%%%%%%%%%%%%%%
%                          RESULTS                         %
%%%%%%%%%%%%%%%%%%%%%%%\section{Results}

@@

\begin{figure}
\center 
{\includegraphics[width=0.98\textwidth]{Figure1/Figure1.pdf}}
\caption{A. Schematic depiction of agent-based model components; blue/red denote tissue cells with/without mutations. Black arrows depict regulation of the cell fate in each cell cycle. Inset depicts major interactions between the immune system and tissue cells.
 %The parameters used in this model can be found in the Supplementary material. Unless otherwise stated, parameter values are assumed to be those found in the Supplement. \tcr{Thoughts on the wording here? Once I have the table, I will use the name of that table.}
B. A representative simulation of one patient. The model parameter values used can be found in Supplementary Table 2. 
The inflammation cycling scheme is represented above the patient dynamics. The vertical dashed line denotes the end of the warmup period. Mut: mutated cells; Mes: mesenchymal cells. 
C. Survival curve for one cohort of patients with the parameter values given in Supplementary Table 2. }
\label{fig:ModelIntro}
\end{figure}

@@

These results represent promising steps in understanding the competing roles of the immune system and EMT during (pre-) development of epithelial cancers, yet much remains to be done. Further development of the inflammation module of this model is important given the large and sometimes paradoxical roles that the inflammatory state exerts on the epithelia and cancer-free survival (Figs. \ref{fig:MOAT} and \ref{fig:VaryINFL_and_MesPars}B-D). Currently, inflammation is modeled as independently cycling between high and low schemes, however many of the agents considered in the model actively contribute to the inflammatory state, thus it would be interesting to consider connecting these components (for example by letting the level of inflammation be conditional on the number of mutant cells in the tissue). Another layer of complexity is revealed by the natural anti-inflammatory role of Tregs. One consequence of the current model is that decreasing the number of Tregs increases the cancer-free survival. Clearly, there exists a trade-off to be accounted for, and adding to the model the main effector function of Tregs could remedy this and add depth to our understanding of the various roles that Tregs play in and around the tumor.
